Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention

### 2017-11-27 ms.tex
why M should be primitive; removed conditional from title; don't talk about modifiers when discussing Ram2012; better single mutation explanation in mutation model 2

% Title page
\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
In their models, the only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
Moreover, their argument was based on the expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots \ge w_n$),
and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible and primitive.

@@

\begin{corollary}[Reproductive value principle]
In the notation of Result~\ref{result:mfp},
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated. Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural evolution.

@@

Eq.~\ref{eq:mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.

>>>>

% Title page
\title{Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
%\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
In their models, the only departure from the \emph{reduction principle} assumptions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant alleles in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
Moreover, their argument was based on the expected increase of the stable population mean fitness, rather than on the invasion success of modifier alleles that modify the mutation rate~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 > w_1 > \ldots > w_n$),
and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
We assume that both the deleterious and the beneficial mutation rates are low, such that two mutations are unlikely to occur in the same individual in one generation: $C_k(\delta + \beta) \ll 1$ for all $k=1, \ldots, n$.
This model has been analyzed by~\citet{Ram2012}.

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible and primitive as long as $\delta, \beta > 0$.

@@

\begin{corollary}[Reproductive value principle]\label{cor:rvp}
In the notation of Result~\ref{result:mfp},
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be primitive.
If $M$ is primitive then individuals of type $k$ can transition into any other type in a finite number of generations.
So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
In contrast, if $M$ is imprimitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population.
For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated. Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural selection.
These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness (e.g. $M$ in Eq.~\ref{eq:model} is primitive).

@@

Eq.~\ref{eq:mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population (see \nameref{sec:mut_model_2}).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and increase the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.
